---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
---

<!--
rmarkdown::render("box-2.Rmd",output_format="word_document")# build website
-->


# Box 5.2 Estimating biomass using a length weight relationship

Commonly a subset of fish captured are measured for length and weight 
and the remaining fish are measured for length only. This is 
advantageous as the variance for total biomass can be reduced by 
accounting for variability in weight among fish by exploiting the 
relationship between length and weight. The length weight relationship 
predicts the expected (i.e., mean) weight of fish where only length was 
measured and the estimate of total biomass then estimated by summing the 
predicted weights. The variance can be estimated as well and 95% 
confidence intervals calculated using the variance of the predicted 
weight. 
 

## Setting up the data

The data is 2 columns, length and weight for common carp in mm and kg 
respectively. There were xx common carp captured. Length was measured on 
all captured common carp and weight was measured for xx fish. 

```{r,eval=FALSE}
ylb_lw<-read.csv("yellow-bass.csv")
# look at the first few rows of the data
head(ylb_lw)
# plot the data

plot(weight_g~length_mm,data=ylb_lw,
    xlab="Length (mm)",
    ylab="Weight (g)")
# add a rug plot to see distribution of lengths
# that do not have lengths to verify adequate coverage
rug(ylb_lw[which(!is.na(ylb_lw$weight_g)),]$length_mm)  
```

## Fitting the length weight relationship

The expected weight for a given length is estimated by log linear 
regression in the R code chunk below. First the length and weight data 
are log transformed. Then a linear model is fit to the log transformed 
data. The data is visualized using the `plot` function. 

```{r}
ylb_lw$ln_length<-log(ylb_lw$length)
ylb_lw$ln_weight<-log(ylb_lw$weight)
plot(ln_weight~ln_length,data=ylb_lw,
    xlab="Log fish length (mm)",
    ylab="Log fish weight (kg)")
    
exp(sum(na.omit(ylb_lw$ln_weight))    )
(sum(na.omit(ylb_lw$weight))    )
    
    sum(na.omit(ylb_lw$ln_weight))
```

A linear regression is fit to the log transformed length and weight data 
using `lm`. And the model output summarized and model fit added to the 
plot of raw data to visualize model fit. 

```{r}
fit<- lm(ln_weight~ln_length,data=ylb_lw)
summary(fit)
plot(ln_weight~ln_length,data=ylb_lw,
    xlab="Log fish length (mm)",
    ylab="Log fish weight (kg)")
abline(coef(fit)[1],coef(fit)[2])
```

The fitted model can be used to calculate the expected weight given the 
length for fish that were not measured for length using the `predict` 
function. Using the argument `se.fit=TRUE` in the predict function 
returns the standard error for each estimated mean weight in the data 
set. The standard error associated with each estimated mean will be used 
to estimate the variance for the biomass estimate. 

```{r}
# calculate the estimated mean weight for a fish of
# a given length
mean_weight<-predict(fit,
    newdata=data.frame(ln_length=ylb_lw$ln_length),
    se.fit = TRUE)
# get the need bits from the list returned
mean_weight<-data.frame(mean_weight=mean_weight$fit,
    std_error=mean_weight$se.fit)

# take a look at the predicted mean weights and standard errors
head(mean_weight)
```

```{r,eval=FALSE}
x<-rlnorm(10,10,0.3)
sd_log_scale= sqrt(var(x))/mean(x)

sd_log_scale*mean=sqrt(var(x))
(sd_log_scale*mean)^2=var(x) # variance for each predication
```




The biomass estimate is now the sum of the mean weights given the 
measured lengths. Recall though weight values remain on log scale.

```{r,eval=FALSE}
ln_biomass<- sum(mean_weight$mean_weight)
```

The variance associated with the biomass estimate is
the sum of the standard errors squared and is calculated
in the R code chunk below. 

```{r,eval=FALSE}
ln_biomass_variance<- sum(mean_weight$std_error)^2
```

The 95% confidence intervals for the biomass estimate are calculated
as previously done for the simple random sample and the stratified random
sample, but in this case the values are still on log scale.

```{r,eval=FALSE}
# t value 
# note the lower value is negative and the upper is positive
t_value<-qt(p=c(0.025,0.975), df=sum(sample_size$weight)-1)
lower_95_ci<- ln_biomass+t_value[1]*sqrt(ln_biomass_variance)
upper_95_ci<- ln_biomass+t_value[2]*sqrt(ln_biomass_variance)
```

The real biomass and 95% confidence interval estimates can be
calculated by exponentiating each value to move from log scale back to 
kilograms.

```{r,eval=FALSE}
# biomass estimate
exp(ln_biomass)
# lower 95% confidence interval for biomass
exp(lower_95_ci)
# upper 95% confidence interval for biomass
exp(upper_95_ci)
```

# References

Colvin, M. E., C. L. Pierce, T. W. Stewart, and S. Grummer. 2012. 
Strategies to control a common carp (_Cyprinus carpio_) population by 
pulsed commercial harvest. North American Journal of Fisheries 
Management 32:1251-1264. 

