---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
---

<!--
rmarkdown::render("box-2.Rmd",output_format="word_document")
-->


# Box 5.2 Estimating biomass using a length weight relationship

Commonly a subset of fish captured are measured for length and weight 
and the remaining fish are measured for length only. This is 
advantageous as the variance for total biomass can be reduced by 
accounting for variability in weight among fish by exploiting the 
relationship between length and weight. The length weight relationship 
predicts the expected weight of fish where only length was measured and 
the estimate of total biomass can be estimated by with a non-parametric 
boostrap that resamples the model residuals to make many new datasets of 
predicted values and then biomass estimated by summing the predicted 
weights. The 95% confidence intervals calculated by finding the 0.025 
and the 0.975 quantiles. However the bias associated with predicting 
weight from length propagates when estimating biomass for a sample and 
results in an underestimate of biomass and therefore some caution is 
needed to account for the bias and some resampling to quantify the
uncertainty in the biomass estimate. This 
box demonstrates: 1) how to fit length weight data and estimate weight, 
albeit biased, 2) apply a bias correction to estimated weights, 3) a 
non-parametric bootstrap approach that shuffles the residuals among 
observed values to estimate 95% confidence intervals for the estimated 
biomass. 

```{r,echo=FALSE}
ylb_lw<-read.csv("yellow-bass.csv")
N<-nrow(ylb_lw)
missing<-sum(is.na(ylb_lw[,2]))
lw<-N-missing
```



# The data

The data is 2 columns, length and weight for yellow bass 
_Morone mississippiensis_ from Clear Lake, IA, in mm and g 
respectively. There were `r N`
 yellow bass in th sample. Length was measured on 
all captured yellow bass and weight was measured for `r lw` fish. To 
estiamte biomass for the sample `r missing`. The code
chunk below reads in the data, looks at the first 6 rows,
and plots the data with a rug plot of the underlying lengths.
The rug plots illustrates where there may be fish that have lengths
but not weight and is useful to ensure the data used to estimate
the weight length relationship covers the lengths that need an 
estimated weight.  



```{r,eval=FALSE}
ylb_lw<-read.csv("yellow-bass.csv")
# look at the first few rows of the data
head(ylb_lw)
# plot the data
plot(weight_g~length_mm,data=ylb_lw,
    xlab="Length (mm)",
    ylab="Weight (g)")
# add a rug plot to see distribution of lengths
# that do not have lengths to verify adequate coverage
rug(ylb_lw[which(!is.na(ylb_lw$weight_g)),]$length_mm)  
```

## Fitting the length weight relationship

The expected weight for a given length is estimated by log linear 
regression in the R code chunk below. First the length and weight data 
are log transformed with the `log()` function. In R the default for the log function is 
the natural log (i.e., $log_{e}$). Then a linear model is fit to the 
log transformed 
data. The linearized data is visualized using the `plot()` function. 

```{r}
ylb_lw$ln_length<-log(ylb_lw$length)
ylb_lw$ln_weight<-log(ylb_lw$weight)
plot(ln_weight~ln_length,data=ylb_lw,
    xlab="Log fish length (mm)",
    ylab="Log fish weight (kg)")
```

A linear regression is fit to the log transformed length and weight data 
using the `lm()` function. And the model output summarized and model fit 
added to the plot of raw data using the `abline()` function to visualize model fit. 



```{r}
# fit the length weight model for log transformed length
# and weight
fit<- lm(ln_weight~ln_length,data=ylb_lw)
# linear model summary information
summary(fit)
# plot the raw log transformed length and weight data
plot(ln_weight~ln_length,data=ylb_lw,
    xlab="Log fish length (mm)",
    ylab="Log fish weight (kg)")
# add a line of predicted weights on log scale
abline(coef(fit)[1],coef(fit)[2])
```

## Bias corrected predicted weight

The variance of the predicted weight is used to correct the bias in the 
predicted weight. To calculate the bias corrected weights for a given 
length the `predict()` function is used to get the uncorrected predicted 
weight.

```{r}
# calculate the estimated mean weight for a fish of
# a given length
ylb_lw$predicted_ln_weight<-predict(fit,
    newdata=data.frame(ln_length=ylb_lw$ln_length))
# take a look at the predicted mean weights and standard errors
head(ylb_lw)
```

The overall model mean squared error is used to peform the bias
correction which is extracted from the fitted `lm()` object
and squared to get the means squared error. 

```{r}
# get mean squared error from model fit
mse<-summary(fit)$sigma^2
```

The next code chunk is a bit complicated but in order to correct the 
predictions the first step is to create the model design matrix and
solve for the inverse. This process needs to be applied to
each predicted. The bias correction is then the predicted
weight on log scale plus the variance divided by 2. 


```{r}
X<-model.matrix(~ln_length,ylb_lw)
X_prime<-t(X)# transpose the design matrix
ylb_lw$var_i<-sapply(1:nrow(ylb_lw),function(i)
    {
    mse*(1+t(as.matrix(X[i,1:2]))%*%(X_prime%*%X)^-1%*%as.matrix(X[i,1:2]))
    })
ylb_lw$W_hat_corrected<- exp(ylb_lw$predicted_ln_weight+ylb_lw$var_i/2)
```

Plotting the corrected and the uncorrected predicted weights
shows that the corrected weight is slighly higher, reflecting 
the bias correction.

```{r}
plot(weight_g~length_mm,ylb_lw)
points(W_hat_corrected~length_mm,ylb_lw,col="red")
# uncorrected predicted weight
ylb_lw$W_hat_uncorrected<-exp(ylb_lw$predicted_ln_weight)
points(W_hat_uncorrected~length_mm,ylb_lw,col="blue")
```

The corrected weight can now be used to estimate biomass 
by summing up the values. Comparing the corrected and the 
uncorrected values, you notice the uncorrected biomass is 
less than the bias corrected biomass. 

```{r}
biomass<-sum(ylb_lw$W_hat_corrected)
biomass_uncorrected<-sum(ylb_lw$W_hat_uncorrected)
```

The log transformation used to linearize the length
weight relationship makes calculating a confidence
interval for the biomass estimate difficult and in 
some cases unreasonable because of the long tails the
occur with a log normal distribution. Bootstrapping
provides a practical way to correct the bias associated
with the log transformation and quantify the uncertainty
around the estimated biomass. The code chunk below
uses the `simpleboot` package to resample the residuals
from the fitted model and the model estimated mean squared
error to estimate unbiased weight and then sum up the 
predictions for 1000 replicates. The 0.025 and the 0.975
quantiles are used for the 95% confidence intervals. 


```{r}
library(simpleboot)
# fit the model, as done above
fit<-lm(ln_weight~ln_length,ylb_lw)
# mean squared error from the fitted model
mse<-summary(fit)$sigma^2 
# the design matrix
X<-model.matrix(~ln_length,ylb_lw)
# transpose of the design matrix
X_prime<-t(X)
# resample the residuals from the fitted model
lboot <- lm.boot(fit, R = 1000,rows=FALSE)
```

The result of the bootstrap is contained in a list
the first element of the list is returned below. 
```{r}
lboot$boot.list[[1]]
```

The `sapply()` function takes each list element
and estimates the predicted weight for each weight,
corrects the bias, and then returns the estimated
biomass for each bootstrap replicate. 
```{r}
boot_biomass<-sapply(lboot$boot.list,function(xx)
    {
    mse<-xx$rstderr^2
    var_i<-sapply(1:nrow(ylb_lw),function(i)
        {
        mse*(1+t(as.matrix(X[i,1:2]))%*%(X_prime%*%X)^-1%*%as.matrix(X[i,1:2]))
        })
        
    return(sum(exp(xx$coef[1]+xx$coef[2]*ylb_lw$ln_length+var_i/2)))
    })
```

The vector of biomass values are plotted below and the 
95% confidence intervals determined using the`quantile()` 
function. The confidence interval is narrow as we would expect given
the relationship of length and weight. 

```{r}
hist(boot_biomass)
abline(v=biomass)
ci<-quantile(boot_biomass,prob=c(0.025,0.975))
ci
```





# References

Colvin, M. E., C. L. Pierce, T. W. Stewart, and S. Grummer. 2012. 
Strategies to control a common carp (_Cyprinus carpio_) population by 
pulsed commercial harvest. North American Journal of Fisheries 
Management 32:1251-1264. 

