---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
bibliography: ref.bib
csl: ecology.csl 
---

<!--
rmarkdown::render("box-6.Rmd",output_format="word_document")# build website
if(m==3){dB<- r*B*(log(Bmax/B)) - C}
if(m==4){dB<- (r/p)*B*(1-(B/Bmax)^p) - C}
if(m==5){dB<- r*B*((Bmax-B)/Bmax) - C}
-->


# Box 6 Fitting a continuous biomass dynamics model to data

## Box overview 

This box demonstrates how to estimate parameters of a continuous 
biomass dynamics model using some simulated harvest data. The box
demonstrates how to:

1. simulate continuous biomass dynamics with harvest and fishery independent monitoring,
1. use grid sampling to identify initial starting values to try fitting the model with, and
1. fit a biomass dynamics model given fishery catch and catch per unit effort from 
fishery independent monitoring


## Libraries needed for analysis

The libraries listed in the code chunk below are required to 
numerically integrate the continuous biomass dynamics models
used in this box exercise. 


```{r}
library(deSolve)
```

## Getting started

Estimating the parameters of a continuous biomass dynamics
model can be done using only harvest data, however there must
be strong contrasts in the biomass dynamics for the estimates
to work. The so called one-way trips described in [@H177] where
stock biomass decreases over time will not provide sufficient 
contrasts (i.e., periods of stock growth and decline in biomass)
to estimate parameters reliably. This box simulates a biomass 
system that undergoes variable harvest as a function of effort and 
has a monitoring program independent of the fishery. 

To perform the analysis you will need to 

1. set up a biomass dynamics function to simulate biomass dynamics given 
harvest and the fishery independent monitoring program, 
2. simulate the catch and fishery independent CPUE,
3. develop a function that returns the sums of squares errors, in this 
case the squared deviations of predicted fishery independent CPUE given 
the fitted model, 
4. evaluate a range of plausible combinations of model parameters to 
identify reasonable initial values, and 
5. use an optimization function to identify the combination of 
parameters that minimizes the sums of squared errors. 



### 1. Setting up the biomass dynamics function

The code below is a function that given the inputs `t`, `x`, and `parms` 
calculates the change in biomass, `dB`, change in harvest `dH`, and the 
fishery independent catch per unit effort `I`. The function below specifies 
that the biomass dynamics conform to the @F75 model of biomass dynamics. New in 
this function relative to the previous boxes is the addition of a fishery independent
monitoring program that is used to estimate relative biomass as CPUE ($I$) is the product
of catchability ($q$) and the biomass ($B$) at time $t$. In the function CPUE is returned
for every timestep. In the function below the CPUE is returned as a separate value after the
state variables (biomass, harvest) as `return(list(c(dB,dH),cpue=I))`. Any value
can be returned from the function in this way but the first item in the list must be
the state variables. 


```{r}
ode_fox<-function(t,x,parms)
    {
    # set the state variable 
    B<-x[1]
    H<-x[2]
    # set the parameters
    r<-parms[1] # intrinsic growth rate
    Bmax<-parms[2] # maximum biomass
    F<-parms[3]    
    q<-parms[4] # catchability for fishery independent sampling
    # effort
    #f<-f_fun(t) # function to 
    # the Fox model of biomass dynamics as a 
    # ordinary differential equation
    ## change in biomass
    dB<- r*B*(log(Bmax/B))  - F*B
    ## change in harvest
    dH<- F*B
    # fishery independent monitoring
    I<- q*B
    return(list(c(dB,dH),cpue=I))# note state variables first in list returned then other values of interest
    }
```

The function is then used by the `ode` function to numerically integrate 
the biomass dynamics (given effort and fishing mortality) and keep track 
of the CPUE for the fishery independent monitoring. 


### Simulate biomass dynamics with varying annual fishing mortality

The code below simulates continuous biomass dynamics given the inputs to 
`ode_fox` specified in the previous section for 20 years. The parameters 
used to govern biomass dynamics are

* $B_{max}$ = 10000,
* $r$ = 0.35,
* $F$ = 0.4, and 
* $q$ = 0.03. 

```{r}
set.seed(12345)
n_years<-20
#f<-runif(n_years,100,5000)
#f_fun<-approxfun(x=c(1:n_years),
#    y=f,
#    method="constant",
#    rule=2)
parameters<-c(r=0.35,
    Bmax=10000,
    F=0.4,
    q=0.03)    
initial_biomass<- c(B=9000,H=0)
solution<- ode(
    y=initial_biomass, 
    times=seq(1,n_years+0.05,by=0.05), 
    func=ode_fox, 
    parms=parameters, 
    method="lsoda")
solution<-as.data.frame(solution)
solution$time<-round(solution$time,2)
plot(solution$B,type="l")
plot(solution$H,type="l")
plot(solution$cpue,type="l")
```


```{r}
solution$year<-floor(solution$time)
# annual catch
catch<-aggregate(cbind(time,H)~year,solution, max)[,c(1,3)]
catch$H<-c(catch$H[1],diff(catch$H))
# CATCH PER UNIT EFFORT 
cpue<- subset(solution,
    time %in% round(seq(1,n_years, 1)+0.5,2))
cpue<-data.frame(time=cpue$time,
    cpue=cpue$cpue.q)
obs<-list(cpue=cpue,catch=catch)
```

The function below contains 3 parts. The function
requires at a minimum an argument to accept the parameter inputs
from the `optim` function `par`. The second argument is 
used to specified the observed yield, needed to calculate
the goodness of fit metric. In the function, the 
first part specifies the inputs needed to project
biomass dynamics from `ode_fox`. The second part
projects the biomass dynamics given the input parameters.
The third part calculates a good of fit metric for each
observation and returns the sum of the metric. 

```{r}
ode_fox<-function(t,x,parms)
    {
    # set the state variable 
    B<-x[1]
    H<-x[2]
    # set the parameters
    r<-parms[1] # intrinsic growth rate
    Bmax<-parms[2] # maximum biomass
    q<-parms[3] # catchability for fishery independent sampling

    # the Fox model of biomass dynamics as a 
    # ordinary differential equation
    ## change in biomass
    catch<- catch_fun(t)
    dB<- r*B*(log(Bmax/B))  - catch
    ## change in harvest
    dH<- catch
    # fishery independent monitoring
    I<- q*B
    return(list(c(dB,dH),cpue=I))
    }
```



```{r}

    # 1. PARAMETER NEED TO PROJECT ode_fox
    ## 1.1. effort 
    catch_fun<-approxfun(x=obs$catch$year,
        y=obs$catch$H,
        method='constant',
        rule=2)

ssq_fun<-function(par,obs)
    {

    # 1.2. BDM PARAMETERS    
    parameters<-c(r=par['r'],   # intrinsic growth rate
        Bmax=par['Bmax'],       # Bmax
        q=par['q'])             # catchability
    ## 1.3. INITIAL BIOMASS AS A FRACTION OF BMAX    
    initial_values<- c(B=par['Bmax']*par['B0'],
        H=0)
    print(parameters)
    # 2. BDM TO BE PROJECTED GIVEN F, BMAX, R, AND B0
    solution<- ode(
        y=initial_values, 
        times=seq(1,n_years+0.05,by=0.05), 
        func=ode_fox, 
        parms=parameters, 
        method="lsoda")
    solution<-as.data.frame(solution)
    
    # 3. OBJECTIVE FUNCTION TO MINIMIZE
    ## 3.1. SUBSET OUT YIELD AT THE END OF THE YEAR
    ##      TO COMPARE TO OBSERVED YIELD
    pred_cpue<- subset(solution,round(time,2)%in% 
        round(obs$cpue$time,2))
    ## 3.2. SUM OF LOG NORMAL MULTPLICATIVE ERRORS
    sserror<-sum((obs$cpue$cpue-pred_cpue$cpue.q)^2)
    sserror<-ifelse(is.na(sserror),100000000,sserror)
    return(sserror) # RETURN OBJECTIVE VALUE
    }
```

Initial starting values are needed for the optimizer.
These values are the best guess at the values. In this 
optimization initial value for B_{max}, $r$, $B_0$ and
$F_{year}$ are needed for the optimization to being. 


```{r}
ini_par<-c(Bmax=10000,
    B0=0.9,
    r=0.35,
    q=0.03)
```

grid search

combos<-expand.grid(
    Bmax=seq(5000,15000,by=500),
    B0=seq(0.5,1,0.1),
    r=seq(0.05,0.5,0.05),
    q=seq(0.005,0.05, 0.005));dim(combos)
out<-lapply(c(1:nrow(combos)),function(x)
    {
    out<- as.data.frame(combos[x,])
    out$ssr<-ssq_fun(par=unlist(combos[x,]), obs=obs) 
    print(x/nrow(combos))
    return(out)
    })
out<-do.call("rbind",out)


The `optim` function is used to return a minimum value, in this case the 
function to be minimized is the sum of squares error returned by 
`ssq_fun`. In the function below the bounded `BFGS` optimizer is used 
because inputs in the model need to be bounded to be greater than 0. The 
upper value bounds are also specified to keep the optimization 
reasonable. In general, if the optimization converges at a boundary then 
you should reevaluate the bounds. If optimization converges within the 
bounds then the solution is likely sufficient, however there are 
benefits to ensuring the solution is the likely the true minimum by 
rerunning the optimization with varying start values to see of the 
optimization results in similar fitted values. 


real sensitive to initial values
ssq_fun(par=ini_par,obs=obs)

```{r}
fit<-optim(par=ini_par,
    fn=ssq_fun,
    method="L-BFGS-B",
    lower=rep(5000,0,0.01,0.001),
    upper=c(70000,1,0.5,0.3),
    control=list(parscale=c(1000,0.1,0.1,0.1)),
    obs=obs)   
options(scipen=15)
fit$par
```



# https://rstudio-pubs-static.s3.amazonaws.com/226974_2b2fded698f448ecaac8f9d5a4480942.html


## References

