---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
---

<!--
rmarkdown::render("box-3.Rmd",output_format="word_document")# build website
-->


# Projecting continous biomass dynamics

This box demonstrates how to project continuous biomass
dynamics from an ordinary differential equation and the
effect of the timestep and numerical integrator.

## Libraries needed for analysis

```{r}
library(deSolve)
```

## Setting up the biomass dynamics model

Projecting biomass dynamics as an ordinary differential equation (ode) or a 
difference model can be done using a function. The function has a 
a couple of parts to it that are common regardless of the model (e.g., Graham Schaefer,
Pella Tomlinson, Fox) or the type of model (ode or difference). The 4 parts within the 
function are:

1. Defining the state variables (e.g., biomass, harvest)
2. Defining the parameters (e.g., intrinsic growth rate, carrying capacity, fishing mortality)
3. Definging the ode or difference equation. This part differs between the 2
types and is demonstrated below. 
4. A `return()` function that is a list of the state variables and other quantities that
are of interest like the amount of fish harvested. 

## Graham Schaefer Model

The function below sets up the Graham Schaefer as an 
ode and that can be used with a numerical integrator from the
`deSolve` package. The function takes 3 input arguments, `t` which 
is the timestep, `x` which is a vector of the state variables, and
`parms` which is a vector of the parameters.

```{r}
ode_gs<-function(t,x,parms)
    {
    # 1. set the state variable 
    B<-x 
    
    # 2. set the parameters
    r<-parms[1]
    K<-parms[2]
    F<-parms[3]
    
    # 3. set up the Graham-Schaefer model of biomass dynamics as an ode
    dB<- r*B*(1-(B/K))-F*B
    
    # 4. return the relevant quantites for the time step
    return(list(dB))
    }
```
To run the model and project biomass the values for each parameter
must be specified in a vector.

```{r}
parameters<-c(r=0.3,K=10000,F=0.1)
```

The initial biomass must also be set to give the prjected biomass dynamics
as starting point. Here the initial biomass is set to 10,000.

```{r}
initial_biomass<- c(B=10000)
```

The `ode()` function from the `deSolve` package provides several 
numerical integrators (e.g., Euler, Runge-Kutta 4)  that can be used to project biomass dynamics. 
There is no set rules for which, in general if biomass dynamics are smooth over time
the type of integrator is a tradeoff of the resolution of the timestep and whether 
dynamics are the same at the end. 

```{r}
solution<- ode(
    y=initial_biomass, 
    times=seq(0,50,by=0.1), 
    func=ode_gs, 
    parms=parameters, 
    method="rk4")
solution<-as.data.frame(solution)
plot(B~time,solution,ylab="Biomass",xlab="Time",las=1,main="")
```



```{r}
ode_gs<-function(t,x,parms)
    {
    # set the state variable 
    B<-x [1]
    Y<-x[2]
    # set the parameters
    r<-parms[1]
    K<-parms[2]
    F<-parms[3]
    
    # the Graham-Schaefer model of biomass dynamics as a 
    # ordinary differential equation
    dB<- r*B*(1-(B/K))-F*B
    # keep track of fishery yield, accumulates over time
    dY<- F*B
    return(list(c(dB,dY)))
    }
parameters<-c(r=0.3,K=10000,F=0.1)
# set initial biomass and yield
initial_biomass<- c(B=10000,Y=0)
solution<- ode(
    y=initial_biomass, 
    times=seq(0,50,by=0.1), 
    func=ode_gs, 
    parms=parameters, 
    method="rk4")
solution<-as.data.frame(solution)
plot(B~time,solution,ylab="Biomass",xlab="Time",las=1,main="",type='l',
    ylim=c(0,10000))
points(Y~time,solution,ylab="Biomass",xlab="Time",las=1,main="")
```


if(m==3){dB<- r*B*(log(Bmax/B)) - C}

### Pella Tomlinson
if(m==4){dB<- (r/p)*B*(1-(B/Bmax)^p) - C}

### Fox
if(m==5){dB<- r*B*((Bmax-B)/Bmax) - C}

## Differential versus difference model

### Numerically integrating continuous biomass dynamics


Keeping track of annual yield

```{r}
ode_gs<-function(t,x,parms)
    {
    # set the state variable 
    B<-x [1]
    Y<-x[2]
    # set the parameters
    r<-parms[1]
    K<-parms[2]
    F<-parms[3]
    
    # the Graham-Schaefer model of biomass dynamics as a 
    # ordinary differential equation
    dB<- r*B*(1-(B/K))-F*B
    # keep track of fishery yield, accumulates over time
    dY<- F*B
    return(list(c(dB,dY),Y=dY))
    }
parameters<-c(r=0.3,K=10000,F=0.1)
# set initial biomass and yield
initial_biomass<- c(B=10000,Y=0)
solution<- ode(
    y=initial_biomass, 
    times=seq(0,50,by=0.1), 
    func=ode_gs, 
    parms=parameters, 
    method="rk4")
solution_cont<-as.data.frame(solution)
plot(B~time,solution_cont,ylab="Biomass",xlab="Time",las=1,main="",type='l',
    ylim=c(0,10000))
points(Y~time,solution_cont,ylab="Biomass",xlab="Time",las=1,main="")
```


## Iterative evaluation of biomass dynamics

Difference model

```{r}
diff_gs<-function(t,x,parms)
    {
    # set the state variable 
    B<-x 
    
    # set the parameters
    r<-parms[1]
    K<-parms[2]
    F<-parms[3]
    
    # the Graham-Schaefer model of biomass dynamics as a 
    # difference equation
    B<-B+r*B*(1-(B/K))-F*B
    return(list(B))
    }
parameters<-c(r=0.3,K=10000,F=0.1)
parameters[1]<-exp(parameters[1])-1
parameters[3]<-exp(parameters[3])-1


initial_biomass<- c(B=10000)
solution<- ode(
    y=initial_biomass, 
    times=seq(0,50,by=0.1), 
    func=diff_gs, 
    parms=parameters, 
    method="iteration")
solution_diff<-as.data.frame(solution)
plot(B~time,solution_diff,ylab="Biomass",xlab="Time",las=1,main="")
points(B~time,solution_cont,col="red")
```







