---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
bibliography: ref.bib
csl: ecology.csl
---

<!--
rmarkdown::render("box-3.Rmd",output_format="all")# build website
-->


# Projecting continuous biomass dynamics

This box demonstrates how to:

1. project continuous biomass dynamics from an ordinary differential equation using R,
2. determine the appropriate time step for the numerical integrator, and
3. the effect of different numerical integrators.

## Libraries needed for analysis

The code below loads the `deSolve` package that can be used to solve
ordinary differential equations (ODEs) or difference equations in 
R [@S242]. 

```{r}
library(deSolve)
```

## Setting up the biomass dynamics model

Projecting biomass dynamics as an ordinary differential equation (ode) 
or a difference model can be done using a function. The function has a 
a couple of parts to it that are common regardless of the model (e.g., 
Graham Schaefer, Pella Tomlinson, Fox) or the type of model (ode or 
difference). The 4 parts within the function are: 

1. Defining the state variables (e.g., biomass, harvest)
2. Defining the parameters (e.g., intrinsic growth rate, carrying capacity, fishing mortality)
3. Definging the ode or difference equation. This part differs between 
the 2 types and is demonstrated below. 
4. A `return()` function that is a list of the state variables and other 
quantities that are of interest like the amount of fish harvested. 
 

## Graham Schaefer Model

The function below sets up the Graham Schaefer model @G108 as an ODE 
where $\frac{dB}{dt}= r\cdot B \cdot \frac{K-B}{K}-F\cdot B$ and that 
can be used with a numerical integrator from the `deSolve` package. The 
function takes 3 input arguments, `t` which is the timestep, `x` which 
is a vector of the state variables, and `parms` which is a vector of the 
parameters. 


```{r}
# Graham Schaefer model specified as an ODE
ode_gs<-function(t,x,parms)
    {
    # 1. set the state variable 
    B<-x 
    
    # 2. set the parameters
    r<-parms["r"]
    K<-parms["K"]
    F<-parms["F"]
    
    # 3. set up the Graham-Schaefer model of biomass dynamics as an ode
    dB<- r*B*(1-(B/K))-F*B
    
    # 4. return the relevant quantities for the time step
    return(list(dB))
    }
```

To run the model and project biomass the values for each parameter must 
be specified in a vector. 


```{r}
parameters<-c(r=0.6,K=10000,F=0.1)
```

The initial biomass must also be set to give the projected biomass 
dynamics as starting point. Here the initial biomass is set to 10,000. 


```{r}
initial_biomass<- c(B=10000)
```


The `ode()` function from the `deSolve` package provides several 
numerical integrators (e.g., Euler, Runge-Kutta 4) that can be used to 
project biomass dynamics. There is no set rules for which, in general if 
biomass dynamics are smooth over time the type of integrator is a 
tradeoff of the resolution of the timestep and whether dynamics are the 
similar throughout the projection. 


```{r}
dt<-c(0.5,0.25, 0.1, 0.01,0.001)
out<-data.frame()
for(i in 1:length(dt))
    {
    solution<- ode(
        y=initial_biomass, 
        times=seq(0,20,by=dt[i]), 
        func=ode_gs, 
        parms=parameters, 
        method="euler")
    solution<-as.data.frame(solution)
    solution$dt<-dt[i]
    out<-rbind(out,solution)
    }
```

The code below plots the solutions for the projected biomass
for decreasing time steps use for the numerical integration. 

```{r}
plot(B~time,solution,
    ylab="Biomass",
    xlab="Time",
    las=1,
    type='n',
    main="")
points(B~time,out,subset=dt==0.5,type="l",lty=1)
points(B~time,out,subset=dt==0.25,type="l",lty=2)
points(B~time,out,subset=dt==0.1,type="l",lty=3)
points(B~time,out,subset=dt==0.01,type="l",lty=4)
points(B~time,out,subset=dt==0.001,type="l",lty=5)
legend("topright",title="Time step",c("0.5","0.25","0.1","0.01","0.001"),
    lty=c(1,2,3,4,5),bty="n")
```

### Interpretation

The ending biomass is the same for all the 5 time steps evaluated. However, there
was some discrepancy in projected biomass up to year 10. The 0.5 time step resulted
in biomass projections that were lower than the 0.01 and 0.001 time steps. The take 
away from this example is that the time step can be important to accurately project
biomass dynamics. As a rule of thumb the smallest time step within the computational
burden allowed should be used. In other words, as time step decreases in size the 
amount of computer time needed to project the dynamics increases, therefore 
pick a time step that is sufficiently accurate for the needs of the analysis and
the model runs in a reasonable amount of time. 

## Comparing numerical integrators

Similar to the issue with time step, the accuracy of numerical integrations
varies with the numerical integrator used. The most basic numerical 
integrator is the Euler (`euler`) while the Runge-Kutta 4 (`rk4`) and 
and the Livermore family (`lsoda`) are much more sophisticated and 
in many cases more accurate at a coarser time step than the Euler [@S233; @S243].
The code chunk below loops over the 3 numerical integrates using the parameters
and initial biomass of the previous example to with a time step of 0.25. 

```{r}
integrator<-c("euler","rk4","lsoda")
out<-data.frame()
for(i in 1:length(integrator))
    {
    solution<- ode(
        y=initial_biomass, 
        times=seq(0,20,by=0.25), 
        func=ode_gs, 
        parms=parameters, 
        method=integrator[i])
    solution<-as.data.frame(solution)
    solution$integrator<-integrator[i]
    out<-rbind(out,solution)
    }
```

The code below plots the solutions for the 3 integrators. 
Similar to the previous results where the `euler` integrator
was used for decreasing time steps, the projections using the
`rk4` and the `lsoda` integrators are identical but slightly
higher than the `euler`. 

```{r}
plot(B~time,solution,
    ylab="Biomass",
    xlab="Time",
    las=1,
    type='n',
    main="")
points(B~time,out,subset=integrator=="euler",type="l",lty=1)
points(B~time,out,subset=integrator=="rk4",type="l",lty=2)
points(B~time,out,subset=integrator=="lsoda",type="l",lty=3)
legend("topright",title="Integrator", c("euler","rk4","lsoda"),
    lty=c(1,2,3),bty="n")
```

## Take home

Any integrator can be used to numerically solve and project
biomass dynamics using ODEs, but sone initial effort should be 
used to evaluate which integrator will be used and what time step
is sufficient for the problem. The most accurate projects will use
small time steps but there are diminishing returns in accuracy
that trade off with model run times. In closing a time step is
sufficiently small if decreasing it does not appreiciable change
the biomass dynamics over time simulated.
