---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
bibliography: ref.bib
csl: ecology.csl 
---

<!--
rmarkdown::render("box-5.Rmd",output_format="word_document")# build website
if(m==3){dB<- r*B*(log(Bmax/B)) - C}
if(m==4){dB<- (r/p)*B*(1-(B/Bmax)^p) - C}
if(m==5){dB<- r*B*((Bmax-B)/Bmax) - C}
-->


# Box 5 Projecting continuous biomass dynamics with pulsed removals

## Box overview 

Inland freshwater fisheries systems are commonly harvested over short 
periods, representing a pulsed biomass removal. This is commonly the 
case when invasive species like common carp _Cyprinus carpio_ and 
bigheaded carp _Hypophthalmichthys nobilis_ use commercial fisheries. 
Additionally, biomass can be removed from a recreational fishery over a 
short period when short seasons occur within an annual time period. 
This box demonstrates how to project semidiscrete biomass dynamics 
models, which are continuous biomass dynamics models that allow for
pulsed dynamics like biomass removals [@C229].


## Libraries needed for analysis

The libraries listed in the code chunk below are required to 
numerically integrate the biomass dynamics models. 


```{r}
library(deSolve)
```

## Projecting semidiscrete biomass dynamics

To project semidiscrete biomass dynamics an additional 
function is needed to simulate the addition or removal 
from the system. First a function that specifies the 
ordinary differential equation the describes the biomass
dynamics is specified in the code chunk below. This box
assumes the Fox biomass dynamics model [@F75]. Note that
in the code chunk, differs from previous box examples as
the change in Yield over each time step is set to 0 because
the amount of biomass removed from the system is a pulsed event
the occurs at specific times. The way that numerical integrators
work, simply specifying the removal amount in the function that 
defines the ODE is not sufficient because the corresponding change 
in biomass and harvest yield is adjusted by the time step ($dt$). 
By fixing `dY` to 0 in the function we can keep track of the amount 
of biomass harvested from the system for evaluating management policies. 


```{r}
ode_fox<-function(t,x,parms)
    {
    # set the state variable 
    B<-x [1]
    Y<-x [2]
    # set the parameters
    r<-parms[1]
    Bmax<-parms[2]

    # the Fox model of biomass dynamics as a 
    # ordinary differential equation
    yield<- F*B
    dB<- r*B*(log(Bmax/B)) 
    dY<- 0
    return(list(c(dB,dY)))
    }
```

To simulate the removal events another function is needed that
is applied as specified time steps. The function has the same
input arguments as `ode_fox`. In the function instantaneous changes
in biomass can be specified. In this example the fishing mortality 
is applied at a single instance and the biomass is reduced by the 
removal amount as `B-B*F`. Additionally, the amount removed is
specified as the biomass times the fishing mortality `B*F`. The
function is set up to keep track of biomass and yield. Defining
the outputs to work with `ode_fox` and solving using `ode` is a 
bit different. The values the function returns needs to be the same
length as the initial values, so a vector `ini` is specified 
below that specifies the initial biomass at `10000` and harvest
yield at `0`. As you will see below this is a bit of a work
around to keep track of the amount of biomass removed. 


```{r}
harvestFun <- function(t, x, parms){
    B<- x[1]
    F<- parms['F']
    x[1] <- B-B*F # biomass - removed biomass
    x[2] <- B*F # biomass removed
    return(x)
    }	
```


Here the model parameters are specified for the intrinsic growth rate
maximum biomass and fishing mortality. A vector of initial biomass
amount is specified and harvest yield as an object `ini`.



```{r}
parms<-c(r=0.3,Bmax=10000,F=0.1)
ini<- c(B=10000,Y=0)
```

The final part of setting up the model to project biomass dynamics with
pulsed removals requires specifying when biomass will be removed and 
the time to evaluate the model. In this example, biomass is pulsed out
of the system at the beginning of year 1, 2, and 4. The model will project the 
biomass dynamics for 10 years using a timestep of 0.1. Experience using events
in the `ode` function had found that at times floating points can result in an 
event not occurring and therefore rounding the times to the corresponding digits
of the timestep minimizes this occurrence.

```{r}
evtimes<-c(1,2,4) # when removals occur
times<-round(seq(0,10,by=0.1),1) # round to time step digits 
```

The model is project using the `ode` function. However a new
argument is included, `events`. The `events` argument can take 
different forms, in this example the function that specifies
how removals occur and the times those events occur are 
specified as a list. 

```{r}
modelout  <-ode(
    y 		= ini, 
    times 	= times, 
    func 	= ode_fox, 
    parms 	= parms,
    events 	= list(func= harvestFun, time=evtimes),
    method 	= "lsoda")
modelout<-as.data.frame(modelout)   
```

The results are plotted using the code chunk below. You can 
see the change in biomass at times 1, 2, and 4 that result from 
the pulsed removal. 

```{r}
plot(B~time,modelout,type='l',ylab="Biomass",xlab="Year",las=1)
```
To fully interpret the amount of biomass removed we need to modify 
the output a bit. In short the model keeps track of biomass and 
harvest yield with changes governed by the ODE. However when removal 
events occur the state variables (biomass, harvest yield) are changed
for that instant. Looking at the results of the output below you can 
see where the harvest event was triggered at time 1, 2, and 4 but the 
harvest yield values then remain the same until the next event. This is 
because the change in harvest yield `dY` was set to 0. The harvest yield is
this rest at the next event. 

```{r}
modelout
```
To clean up this artifact the harvest yield at times other than when 
the event occurred are set to 0 using the code below. Note that 
while the event was triggered at time 1, 2, and 4 the harvest yield is 
recorded in timestep 1.1, 2.1, and 4.1. The code below sets values
other than the event times (i.e., 1, 2, 4) plus the time step to
0 as harvest did not occur at those times, but the `ode` function 
carried the harvest yield over those steps because `dY` was set
to 0 so that the harvest yield could be tracked in the projection.
The `round` function is used to ensure the right times are selected. 

```{r}
modelout[!(modelout$time%in%(round(evtimes+0.1,1))),]$Y<-0
```
The plot below illustrates the pulsed harvest removals that occurred at time
1, 2, and 4. 

```{r}
plot(Y~time,modelout,type="h",ylab="Harvest",xlab="Year",las=1)
```

## References

