---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
---

<!--
rmarkdown::render("box-4.Rmd",output_format="word_document")
-->


# Projecting biomass dynamics to find biological reference points

## Box overview 

Analytical solutions exist to determine biological reference points like
$F_{MSY}$, $B_{MSY}$, and $MSY$
for biomass dynamics models like the Graham Schaefer, Fox, and the Pella
Tomlinson. However there are instances where analytical solutions might not
be available because the underlying biomass dynamics model was modified to 
account for more biological realism. Additionally, the analytical solutions
for biological reference points assume continuous biomass dynamics and therefore
if a difference equation is used to project biomass dynamics, biological reference points
may be biased. This box demonstrates how to find 
the biological reference points for a biomass dynamics model by simulation of 
a biomass dynamics model implemented as an ordinary differential equation or
as a difference equation. 

This box uses the Fox biomass dynamics model to project biomass dynamics over 
100 years for varying fishing mortality rates ($F$). The time horizon of 100 years
was sufficient for this example for the biomass dynamics to reach equilibrium (i.e., 
change in biomass from one time point to the next is ~0). There are 2 examples, one 
for the Fox model implemented as an ODE and another with the Fox model implemented as 
a difference model. 

## Libraries needed for analysis

The libraries listed in the code chunk below are required to plot the 
biomass dynamics and numerically integrate the biomass dynamics models. 

```{r}
library(lattice)
library(deSolve)
```

## Evaluated reference points for a continuous biomass dynamics model

The function below sets up the Fox model of biomass dynamics as an 
ODE.

```{r}
ode_fox<-function(t,x,parms)
    {
    # set the state variable 
    B<-x 
    
    # set the parameters
    r<-parms[1]
    Bmax<-parms[2]
    F<-parms[3]
    
    # the Fox model of biomass dynamics as a 
    # ordinary differential equation
    yield<- F*B
    dB<- r*B*(log(Bmax/B))  - F*B
    return(list(c(dB),Y=as.vector(yield)))# as.vector drops Y.F name baggage
    }
```

The code chunk below sets the parameters of the model, initial biomass,
and the then numerically integrates the model using the `lsoda` numerical 
integration routine. The projection biomass dynamics are saved as a `data.frame`.

```{r}
parameters<-c(r=0.3,Bmax=10000,F=0.1)
initial_biomass<- c(B=10000)
solution<- ode(
    y=initial_biomass, 
    times=seq(0,100,by=0.1), 
    func=ode_fox, 
    parms=parameters, 
    method="lsoda")
solution<-as.data.frame(solution)
```

The code chunk below plots the biomass dynamics. Looking at the 
biomass dynamics approaching year 100 there is no visible change in
biomass, suggesting equilibrium has been achieved. 

```{r}
plot(B~time,solution,ylab="Biomass",
    xlab="Time",las=1,main="",type="l")
```

The code chunk below returns the last 6 rows of the solution and
the yield or biomass values do not change appreciably confirming that
the biomass dynamics are in what is called approximate equilibrium (i.e., 
changes in biomass dynamics are too small to be consequential). 

```{r}
tail(solution)
```

To evaluate biological reference points the biomass and yield at 
approximate equilibrium need to be identified for $F$ varying from 0 to 
a value that that exceeds the level that overfishing occurs so that the 
maximum sustained yield can be identified. This is don be looping over
$F$ values and projecting the biomass dynamics out for a sufficiently long 
time, 100 years in this example. The code chunk below performs the loop
and numerical integration. The results are collected in a `data.frame` 
using the `rbind` function for each value of $F$ evaluated.  

```{r}
F_eval<-seq(0,1.6,by=0.01)
out<-data.frame()
for(i in 1:length(F_eval))
    {
    parameters<-c(r=0.3,Bmax=10000,F=F_eval[i])
    initial_biomass<- c(B=10000)
    solution<- ode(
        y=initial_biomass, 
        times=seq(0,100,by=0.1), 
        func=ode_fox, 
        parms=parameters, 
        method="rk4")
    solution<-as.data.frame(solution)
    solution$F<- F_eval[i]
    out<-rbind(out,solution)
    }
```

The code chunk below plots the biomass dynamics for each $F$ value
evaluated for visual evaluation to ensure approximate equlibrium was 
reached for all $F$ values. 

```{r}
xyplot(B~time,out,group=F,type="l",xlab="Time",
    ylab="Biomass")   
```

A quick look at the last 6 rows of the dataset we can see that 
the last time step evaluated is 100. We can use the `subset` function to extract the 
last value of biomass and yield for each $F$ value 
evaluated. 

```{r}
tail(out)
equilibrium_values<- subset(out, time==100)
```

Plotting yield versus $F$ illustrates the sustained or 
equilibrium yield curve. The top of the curve is the 
maximum sustained yield ($MSY$).

```{r}
# plot equilibrium yield values
plot(Y~F,equilibrium_values,ylab="Sustained yield",
    xlab="Time",las=1,main="",
    type='l')
```

## Biological reference points

Biological reference points can be calculated from the equilibrium 
yield curve. Specifically, the biomass and $F$ values associated with the $MSY$ can be
identified by using the `which.max` function to return the row of a `data.frame` 
that has the maximum of a vector. 

```{r}
# get biomass and F that maximize equilibrium yield
bio_ref_points<-equilibrium_values[which.max(equilibrium_values$Y),]
```

The value contained in `bio_ref_points` are $B_{MSY}$ and $F_{MSY}$. 

```{r}
bio_ref_points
```


## Finding $F_{0.1}$

The numerical approach can also be used to identify the more conservative
$F_{0.1}$. The code chunk below calculates the slope over the equilibrium yield 
curve. The slope at the origin is then multiplied by 0.1 and the tangent
line on the curve identified. 

```{r}
equilibrium_values$slopes<- c(0,diff(equilibrium_values$Y)/0.01)
slope10<-equilibrium_values$slopes[2]*0.1
```

The `approxfun` function is useful here to interpolate values. The 
code chunk below sets up an interpolation function that returns an
$F$ and yield value associated with a slope. The code chunk below
returns the $F$ and yield associated with 10% of the yield slope at the origin. 

```{r}
find_F<- approxfun(equilibrium_values$slopes,equilibrium_values$F)
f_01<-find_F(slope10)
find_Y<- approxfun(equilibrium_values$slopes,equilibrium_values$Y)
Y_01<-find_Y(slope10) 
```

The plot below shows the 10% of the slope at the origin, the 
$F$ value that maximizes sustained yield ($F_{MSY}$) and $F$ where
the slope of the tangent line of sustained yield curve is 10% of the slope 
of a tangent at the origin. 

```{r}
plot(Y~F,equilibrium_values,ylab="Sustained yield",
    xlab="Fising mortality",las=1,main="",
    type="l")
abline(0,slope10,lwd=3,lty=2)
points(Y~F,bio_ref_points,pch=19)
points(f_01,Y_01,pch=17)
legend("bottomright", lty=c(2,NA,NA),pch=c(NA,19,17),
    legend=c("Slope at the origin",expression("F"[MSY]), expression("F"[0.1])))
```



## Evaluated reference points for a discrete biomass dynamics model

The code chunk below replicates the analysis done above but for 
the Fox model implemented as a difference model assuming a discrete
time step of 1 year. 


```{r}
diff_fox<-function(t,x,parms)
    {
    # set the state variable 
    B<-x 
    
    # set the parameters
    r<-parms[1]
    Bmax<-parms[2]
    F<-parms[3]
    
    # the Fox model of biomass dynamics as a 
    # difference equation
    yield<- F*B
    B<-B + r*B*(log(Bmax/B))  - F*B
    return(list(c(B),Y=as.vector(yield)))# as.vector drops Y.F name baggage
    }
# evaluate sustained yield
F_eval<-seq(0,1,by=0.01)
out<-data.frame()
for(i in 1:length(F_eval))
    {
    parameters<-c(r=0.3,Bmax=10000,F=F_eval[i])
    initial_biomass<- c(B=10000)
    solution<- ode(
        y=initial_biomass, 
        times=seq(0,100,by=1), 
        func=diff_fox, 
        parms=parameters, 
        method="iteration")
    solution<-as.data.frame(solution)
    solution$F<- F_eval[i]
    out<-rbind(out,solution)
    }
# look at sustained yield by F
xyplot(B~time,out,group=F,type="l",xlab="Time",
    ylab="Biomass")   
# subset values at year 100
equilibrium_values<- subset(out, time>99.9999)
# plot equilibrium values
plot(Y~F,equilibrium_values,ylab="Sustained yield",
    xlab="Fishing mortality",las=1,main="",
    type="l")
# subset biological reference points
bio_ref_points<-equilibrium_values[which.max(equilibrium_values$Y),]
bio_ref_points
# find F_0.1
equilibrium_values$slopes<- c(0,diff(equilibrium_values$Y)/0.01)
slope10<-equilibrium_values$slopes[2]*0.1
find_F<- approxfun(equilibrium_values$slopes,equilibrium_values$F)
f_01<-find_F(slope10)
find_Y<- approxfun(equilibrium_values$slopes,equilibrium_values$Y)
Y_01<-find_Y(slope10)
# illustrate F_MSY and F_0.1
plot(Y~F,equilibrium_values,ylab="Sustained yield",
    xlab="Fishing mortality",las=1,main="",
    type="l")
points(Y~F,bio_ref_points,pch=19)
points(f_01,Y_01,pch=17)
legend("bottomright", pch=c(19,17),
    legend=c(expression("F"[MSY]), expression("F"[0.1])))
```









