---
title: ""
header-includes:
  - \usepackage{amsmath, graphicx}
output: 
  html_document:
    theme: readable
    highlight: tango
  word_document:  
    fig_width: 6.5  
    fig_height: 6.5 
    reference_docx: markdown_word_template_1.docx
---

<!--
rmarkdown::render("box-1.Rmd",output_format="word_document")# build website
-->


# Box 5.1. Estimating biomass from a random sample and a stratified random sample


This box demonstrates how to estimate biomass,associated uncertainties, 
and 95% confidence intervals for a simple random sample, a stratified 
random sample, and for length weight data. Common to the 3 approaches is 
the use of a sampling frame (how many fish are there) and random sample 
of fish from the sampling frame. In short, the mean weight of fish in 
the sample is multiplied by the total number of fish to estimate biomass 
where the mean can be calculated as a simple random sample, from a 
stratified sample, or accounting for variation among all the fish in the 
sampling frame with a covariate like fish length. 

The data used in this example is from commercial common carp _Cyprinus 
carpio_ harvest in Clear Lake, Iowa (Colvin 2012). Commercial harvesters 
capture common carp in the spring and fall annually to remove biomass 
from the system. The biomass removed from the system can be estimated 
using the 2 approaches. The sampling frame is the number of common carp 
harvested for each occasion. The sampling unit is individual common 
carp. 



# Estimating biomass for a simple random sample

## Setting up the data

The R code chunk below sets up the sampling frame and the data for the 
simple random sample. 

```{r}
# weight data in grams for the
# simple random sample
carp_weights<-data.frame(season=c(rep("spring", 180)),
    weight=c(0.611, 0.595, 0.601, 0.629, 0.63, 
    0.635, 0.65, 0.65, 0.65, 0.654, 0.63, 0.64, 0.67, 0.63, 0.648, 0.65, 
    0.634, 0.671, 0.759, 0.645, 0.645, 0.701, 0.66, 0.66, 0.666, 0.669, 
    0.674, 0.675, 0.677, 0.678, 0.68, 0.681, 0.71, 0.654, 0.662, 0.665, 
    0.671, 0.68, 0.682, 0.682, 0.633, 0.662, 0.665, 0.675, 0.65, 0.667, 
    0.681, 0.674, 0.674, 0.681, 0.706, 0.71, 0.698, 0.724, 0.663, 0.674, 
    0.694, 0.71, 0.72, 0.682, 0.685, 0.708, 0.692, 0.707, 0.712, 0.685, 
    0.687, 0.69, 0.702, 0.705, 0.73, 0.685, 0.694, 0.724, 0.725, 0.735, 
    0.745, 0.704, 0.715, 0.691, 0.715, 0.731, 0.747, 0.72, 0.725, 0.71, 
    0.713, 0.729, 0.73, 0.73, 0.705, 0.722, 0.728, 0.735, 0.731, 0.749, 
    0.776, 0.705, 0.751, 0.772, 0.754, 0.757, 0.76, 0.76, 0.79, 0.725, 
    0.739, 0.742, 0.743, 0.753, 0.755, 0.76, 0.765, 0.74, 0.745, 0.731, 
    0.745, 0.78, 0.72, 0.745, 0.755, 0.767, 0.78, 0.768, 0.79, 0.715, 0.766, 
    0.77, 0.79, 0.8, 0.823, 0.751, 0.779, 0.78, 0.795, 0.75, 0.766, 0.786, 
    0.684, 0.75, 0.757, 0.765, 0.77, 0.783, 0.79, 0.785, 0.799, 0.8, 0.76, 
    0.78, 0.81, 0.79, 0.77, 0.79, 0.78, 0.79, 0.79, 0.77, 0.797, 0.808, 
    0.819, 0.805, 0.805, 0.83, 0.82, 0.838, 0.846, 0.83, 0.818, 0.83, 0.814, 
    0.82, 0.821, 0.852, 0.796, 0.84, 0.833, 0.862, 0.86, 0.859)) 

sample_size<-length(carp_weights$weight)
# specify the sampling frame
# 3367 common carp captured but not weighed
# therefore the sampling frame is 2877+ plus the number of carped weighed
sampling_frame<- 3367+nrow(carp_weights)
```

## Estimating the mean and variance of the sample

The R code chunk below estimates the mean and variance of the simple 
random sample 

```{r}
# estimate the mean weight for the sample
mean_weight<-mean(carp_weights$weight)
# estimate the variance of the sample
variance_weight<- var(carp_weights$weight)
```

## Estimating total biomass 

The R code chunk below estimates the total biomass of the common carp 
captured as the product of the mean weight and the number of fish 
harvested (i.e., sampling frame). 

```{r}
biomass<- sampling_frame*mean_weight
# return the biomass estimate
biomass
```

The R code chunk below uses the variance equation provided in table 5.1 
to estimate the variance for the estimated total biomass. The standard 
deviation is calculated as the square root of the variance. 


 
 
```{r}
# estimated variance for biomass estimate
variance_biomass<- sampling_frame^2*(variance_weight/sample_size)
variance_biomass
```

The 95% confidence intervals are calculated by adding the appropriate 
t-value given the sample size as the degrees of freedom. For a 95% 
confidence interval the t-value is calculated using the `qt` function 
given probabilities of 0.025 and 0.975. The probabilities for the lower 
and upper are specified and the degrees of freedom is the sample size. 
The lower and upper 95% confidence interval are calculated by adding the 
product of the lower and upper t value and the standard deviation for 
the total biomass estimate. 



```{r}
# t-value 
# note the lower value is negative and the upper is positive
t_value<-qt(p=c(0.025,0.975), df=sample_size)
# calculate the lower and upper 95% confidence interval for
# the biomass estimate
lower_95_ci<- biomass+t_value[1]*sqrt(variance_biomass)
upper_95_ci<- biomass+t_value[2]*sqrt(variance_biomass)
```

The values below are the lower 95% confidence interval, the total 
biomass and the upper 95% confidence interval respectively. 

 

```{r}
lower_95_ci
biomass
upper_95_ci
```


# Estimating biomass for a stratified random sample

There are instances where stratification is necessary to estimate 
biomass. Continuing with the common carp harvest from Clear Lake, Iowa 
there were harvest events that occurred in the spring and the fall. It 
makes sense for annual reporting to report the estimated total biomass 
removed but there are differences in fish weight that should be 
accounted for in the estimate. Specifically, fish in the spring, after 
ice out, tend weigh less than fish in the fall after a season of growth. 
Therefore the time of harvest (i.e., spring, autumn) can be used to 
account for the weight differences that effect sample. Similar to the 
simple random sample the data required is a sampling frame and the 
sample random sample within the frame, but for each strata. For this 
example there are 2 strata, spring and autumn. In application this type 
of estimate can be used if there are weight differences between sex and 
among locations or life stages. 

## Setting up the data

The R code chunk below sets up the sampling frame and the data for the 
stratified random sample. The stratum weight is calculated
because it is need to calculate the weighted mean fish weight. The 
stratum weight is the number of sampling units in each stratum divided
by the total number of sampling units. 

```{r}
# weight data in kilograms for the carp harvested in the autumn
# 41 fish with weights and another 743 without weights
carp_weights<-rbind(carp_weights,data.frame(season=c(rep("autumn", 41)),
    weight=c(0.652, 0.632, 0.668, 0.684, 0.69, 0.695, 0.703, 0.698, 0.692, 
        0.714, 0.695, 0.74, 0.715, 0.729, 0.742, 0.713, 0.715, 0.745, 0.759, 
        0.761, 0.768, 0.725, 0.809, 0.775, 0.779, 0.798, 0.813, 0.795, 0.811, 
        0.811, 0.79, 0.801, 0.814, 0.835, 0.834, 0.83, 0.82, 0.853, 0.88, 0.88, 
        0.915)) )

# specify the sampling frame
sampling_frame<- data.frame(
    season=c("spring","autumn"),
    N=c(3367+180,743+41))
# calculate the stratum weight, do not confuse with fish weight
sampling_frame$stratum_weight<- sampling_frame$N/sum(sampling_frame$N)
```

## Estimating the mean and variance of the sample

The R code chunk below estimates the mean and variance of
the simple random sample taken within each stratum using
the `aggregate` function. 

```{r}
# estimate the mean weight for each stratum
mean_weight<-aggregate(weight~season,
    data=carp_weights,
    FUN=mean)
sample_size<-aggregate(weight~season,
    data=carp_weights,
    FUN=length)
# estimate the variance for each stratum
variance_weight<- aggregate(weight~season,
    data=carp_weights,
    FUN=var)
# return the stratum specific mean fish weight
mean_weight
# return the stratum specific variance
variance_weight
```

## Estimating total biomass for the stratified sample

The R code chunk below estimates the weighted mean 
of the common carp captured in the spring and the autumn. 
Noe that `sampling_frame` is not in the same order as 
the `mean_weight` and `variance_weight`. The `order` function
sorts the `data.frame` so it matches the output of the
`aggregate` function. 

```{r}
sampling_frame<-sampling_frame[order(sampling_frame$season),]
weighted_mean_weight<- mean_weight$weight*sampling_frame$stratum_weight
```

The R code chunk below calculates the overall mean fish weight. 
The biomass is then estimated as the total number of sampling 
units times the overall mean fish weight. 

```{r}
# overall mean fish weight
mean_weight<- sum(weighted_mean_weight)
# biomass estimate
biomass<- sum(sampling_frame$N)*mean_weight
biomass
```

The estimated variance for the biomass estimate is the sum of 
the stratum specific variance estimates. The stratum specific
variance is the same as the biomass variance for the simple
random sample, but in the case of a stratified sample applied
to each stratum. The standard deviation is calculated
as the square root of the variance and used in the 
calculation of 95% confidence intervals.  


```{r}
# estimated variance for biomass estimate
variance_biomass<- sum(sum(sampling_frame$N)^2*variance_weight$weight/sample_size$weight)
```

The 95% confidence intervals are calculated using the biomass
estimate, a t-value, and the standard deviation as was done
for the simple random sample. 


```{r}
# t value 
# note the lower value is negative and the upper is positive
t_value<-qt(p=c(0.025,0.975), df=sum(sample_size$weight)-1)
lower_95_ci<- biomass+t_value[1]*sqrt(variance_biomass)
upper_95_ci<- biomass+t_value[2]*sqrt(variance_biomass)
```

The values below are the lower 95% confidence interval, the total 
biomass and the upper 95% confidence interval respectively. 

```{r}
lower_95_ci
biomass
upper_95_ci
```

# References

Colvin, M. E., C. L. Pierce, T. W. Stewart, and S. Grummer. 2012. 
Strategies to control a common carp (_Cyprinus carpio_) population by 
pulsed commercial harvest. North American Journal of Fisheries 
Management 32:1251-1264. 

